---
title: "BMI 714 Final Project"
author: "Wenzhu Ye, Lindsay Cheng"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_engines$set(txt = function(options) {
  code <- paste(options$code, collapse = "\n")
  knitr::engine_output(options, code, NULL)
})
```

```{r}
# Load Library here
library(dplyr)
library(ggplot2)
```

# View Data

```{r}
nhanes <- read.csv("data/BMI714_NHANES2020_Data.csv", header = T)
nhanes_var_name <- read.csv("data/BMI714_NHANES_VariableDictionary.csv", header = T)
# Rename the columns in nhanes
colnames(nhanes) <- ifelse(colnames(nhanes) %in% nhanes_var_name$BMI_714_Variable_Name, 
                           nhanes_var_name$Variable_Name[match(colnames(nhanes),nhanes_var_name$BMI_714_Variable_Name)], 
                           colnames(nhanes))  # Keep names unchanged if not in mapping


nhanes %>% nrow()
```


```{r}
nhanes_var_name %>% select(Data_File_Description) %>% unique()
```


weight in kg: BMXWT
```{r}
nhanes %>% select(BMXWT) %>% is.na() %>% sum()
```


## independent variable consideration
Dairy porduct used: Greater consumption of total dairy products prevention of weight gain in middle-aged and elderly women who are initially normal weight. https://pmc.ncbi.nlm.nih.gov/articles/PMC4807700/

Type of Milk: 1%/skim milk is more common among overweight/obese preschoolers, potentially reflecting the choice of parents to give overweight/obese children low-fat milk to drink. Nevertheless, 1%/skim milk does not appear to restrain body weight gain between 2–4 years in this age range, emphasizing a need for weight-targeted recommendations with a greater evidence basis. https://pmc.ncbi.nlm.nih.gov/articles/PMC4439101/



