---
title: "BMI 714 Final Project"
author: "Wenzhu Ye, Lindsay Cheng"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_engines$set(txt = function(options) {
  code <- paste(options$code, collapse = "\n")
  knitr::engine_output(options, code, NULL)
})
```

```{r}
# Load Library here
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(ggcorrplot)
```

# View Data

```{r}
nhanes <- read.csv("data/BMI714_NHANES2020_Data.csv", header = T)
nhanes_var_name <- read.csv("data/BMI714_NHANES_VariableDictionary.csv", header = T)
# Rename the columns in nhanes
colnames(nhanes) <- ifelse(colnames(nhanes) %in% nhanes_var_name$BMI_714_Variable_Name, 
                           nhanes_var_name$Variable_Name[match(colnames(nhanes),nhanes_var_name$BMI_714_Variable_Name)], 
                           colnames(nhanes))  # Keep names unchanged if not in mapping


nhanes %>% nrow()
```
```{r}
nhanes %>% select(RIDAGEYR) %>% filter(RIDAGEYR >= 18) %>% nrow()
```

```{r}
nhanes_var_name %>% select(Data_File_Description) %>% unique()
```


weight in kg: BMXWT
```{r}
# Calculate the percentage of missing values for each column
missing_percent <- colSums(is.na(nhanes)) / nrow(nhanes)

# Filter columns with less than 10% missingness
filtered_df <- nhanes[, missing_percent < 0.20]
dim(filtered_df)
```

```{r}
# Extract the column names from the filtered dataset
filtered_columns <- colnames(filtered_df)

# Filter var_dict for rows where col_name matches the filtered column names
filtered_var_dict <- nhanes_var_name[nhanes_var_name$Variable_Name %in% filtered_columns, ]
```

```{r}
nhanes %>% 
  select(ALQ111, RIDAGEYR) %>% 
  filter(RIDAGEYR > 18) %>%
  filter(rowSums(is.na(.)) == 1) %>%  # Check if all selected columns are NA
  nrow()
```

```{r}
nhanes %>% 
  select(ALQ111, RIDAGEYR) %>% 
  filter(RIDAGEYR > 18)
```


```{r}
nhanes %>% select(BMXWT) %>% is.na() %>% sum()
```
## independent variable consideration
variable details in nhanes website: https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?Cycle=2017-2020

Dairy product used(DBQ197: https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/P_DBQ.htm#DBQ197): Greater consumption of total dairy products prevention of weight gain in middle-aged and elderly women who are initially normal weight. https://pmc.ncbi.nlm.nih.gov/articles/PMC4807700/

Type of Milk(DBQ223A, DBQ223B, DBQ223C, DBQ223D, DBQ223E, DBQ223U: https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/P_DBQ.htm#DBQ223A): 1%/skim milk is more common among overweight/obese preschoolers, potentially reflecting the choice of parents to give overweight/obese children low-fat milk to drink. Nevertheless, 1%/skim milk does not appear to restrain body weight gain between 2–4 years in this age range, emphasizing a need for weight-targeted recommendations with a greater evidence basis. https://pmc.ncbi.nlm.nih.gov/articles/PMC4439101/
Missingness:
```{r}
(nhanes %>% 
  select(DBQ223A, DBQ223B, DBQ223C, DBQ223D, DBQ223E, DBQ223U) %>% 
  filter(rowSums(is.na(.)) == ncol(.)) %>%  # Check if all selected columns are NA
  nrow()) / nrow(nhanes) # Count the number of such rows
```


Caffeine: caffeine with weight loss: https://pubmed.ncbi.nlm.nih.gov/30335479/
 - high missingess: 15538/15560
```{r}
caffeine_rows <- nhanes_var_name[grep("caffeine", nhanes_var_name$Variable_Description, ignore.case = TRUE), ]
```

Income(INDFMMPC: https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/P_INQ.htm#INDFMMPC): https://pubmed.ncbi.nlm.nih.gov/29306894/

Sleep(SLQ300, SLQ310, SLQ320, SLQ330: https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/P_SLQ.htm#SLD012):https://pmc.ncbi.nlm.nih.gov/articles/PMC9031614/

Smoking(SMD460: https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/P_SMQFAM.htm#SMD460): https://pubmed.ncbi.nlm.nih.gov/7039293/

Diabetes(DIQ010: https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/P_DIQ.htm#DIQ010): https://www.baptisthealth.com/blog/family-health/does-diabetes-cause-weight-loss-or-weight-gain

Frozen meals/ready to go meal/fast food/resturation: (DBD905, DBD910: https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/P_DBQ.htm#DBD905): https://pmc.ncbi.nlm.nih.gov/articles/PMC4302389/

race(RIDRETH3, https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/P_DEMO.htm#RIDRETH1): https://pmc.ncbi.nlm.nih.gov/articles/PMC3464818/

Alcohol consumption(ALQ121: https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/P_ALQ.htm#ALQ121): 
- note: the data only ask this question for people 18 age and up. Should we classify people below don't consumption any alcohol?
Missing: 
```{r}
(nhanes %>% 
  select(ALQ111, RIDAGEYR) %>% 
  filter(RIDAGEYR > 18) %>%
  filter(rowSums(is.na(.)) == 1) %>%
  nrow()) / (
nhanes %>% 
  select(ALQ111, RIDAGEYR) %>% 
  filter(RIDAGEYR > 18) %>% nrow())
```


protein(DR2TPROT) , carb (DR2TCARB), fat (DR2TTFAT) intake: 
 - https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/P_DR2TOT.htm#DR2TPROT
 - can have more specific group
 - 每个都是3673/14300 missing
 - continuous variable
 - present in doc not in the dataset?

```{r}
nhanes %>% select(RIDAGEYR, DS1TCAFF) %>% filter(RIDAGEYR >= 18) %>% select(DS1TCAFF) %>% is.na() %>% sum()
```

# Variable Choosen: 
Dependent variable: 
weight(BMXWT)

Independent variable:
Dairy product used(DBQ197)
Type of Milk(DBQ223A, DBQ223B, DBQ223C, DBQ223D, DBQ223E, DBQ223U)
Frozen meals/ready to go meal/fast food/resturation (DBD905, DBD910)
Income(INDFMMPC)
Sleep(SLQ300, SLQ310, SLQ320, SLQ330)
Smoking(SMD460)
Diabetes(DIQ010)
Alcohol consumption(ALQ121)

Goal: how socioeconmic state (income), eating behaviour affect obesity (BMI) in adult

# Prep data
```{r}
data <- nhanes %>% 
  select(	
BMXBMI, DBQ197, DBQ223A, DBQ223B, DBQ223C, DBQ223D, DBQ223E, DBQ223U, DBD905, DBD910, INDFMMPC, SLQ300, SLQ310, SLQ320, SLQ330, SMD460, DIQ010, ALQ121, RIDAGEYR) %>% # select variables
  filter(!is.na(	
BMXBMI)) %>% # remove rows with missing weight
filter(RIDAGEYR >= 18) # remove rows with age < 16)

```

## BMI category
according to https://www.cdc.gov/bmi/adult-calculator/bmi-categories.html
```{r}
data <- data %>%
  mutate(
    bmi_category = case_when(
      BMXBMI < 18.5 ~ "Underweight",
      BMXBMI >= 18.5 & BMXBMI < 25 ~ "Normal",
      BMXBMI >= 25 & BMXBMI < 30 ~ "Overweight",
      BMXBMI >= 30 ~ "Obesity",
      TRUE ~ NA_character_  # Handle missing or unexpected BMI values
    )
  )
```


## Process independent variable
Note: predictors are ordinal, encoding them to reflect their order.
```{r}
########### Type of Milk ###########
# For Type of Milk DBQ223A, DBQ223B, DBQ223C, DBQ223D, DBQ223E, DBQ223U, we will combine them into one variable. 
#  want to see combined effect of drinking mupltiple type of milk: Group participants who drink multiple types of milk into a single “Mixed Milk” category.
data <- data %>%
  rowwise() %>%
  mutate(milk_type = if_else(sum(!is.na(c(DBQ223A, DBQ223B, DBQ223C, DBQ223D, DBQ223E, DBQ223U))) > 1, 
                             "Mixed Milk", 
                             as.character(coalesce(DBQ223A, DBQ223B, DBQ223C, DBQ223D, DBQ223E, DBQ223U)))) %>%
  ungroup()

# rename milk_type
data <- data %>% mutate(milk_type = case_when(
  milk_type == "10" ~ '4', #Whole
  milk_type == "11" ~ '3', #2%
  milk_type == "12" ~ '1', # 1% or 0.5%
  milk_type == "13" ~ '0', #Skim
  milk_type == "14" ~ '2', #Soy, ~ 1.5%-2% 
  milk_type == "30" ~ '8', #Other
  milk_type == "Mixed Milk" ~ '9', #Mixed Milk
  milk_type == "77" ~ NA_character_, #Refused
  milk_type == "99" ~ NA_character_, #Unknown
  TRUE ~ NA_character_ # NA into Unknown
))


########### Milk consumption  ###########
data <- data %>%
  mutate(
    milk_consumption = case_when(
      DBQ197 == 0 ~ '0', #Never
      DBQ197 == 1 ~ '1', #less than once a week
      DBQ197 == 2 ~ '2', #once a week or more, but less than once a day
      DBQ197 == 3 ~ '3', # once a day or more
      DBQ197 == 4 ~ '9', #Varied
      DBQ197 == 7 ~ NA_character_, #Refused
      DBQ197 == 9 ~ NA_character_, #Unknow
      is.na(DBQ197) ~ NA_character_  # Handle missing values
    )
  ) 


########### Poverty  ###########
data <- data %>%
  mutate(
    poverty_category = case_when(
      INDFMMPC == 1 ~ '1', #Monthly poverty level index <= 1.30
      INDFMMPC == 2 ~ '2',# 1.30 < Monthly poverty level index <= 1.85
      INDFMMPC == 3 ~ '3', #> 1.85
      INDFMMPC == 7 ~ NA_character_, #Refused
      INDFMMPC == 9 ~ NA_character_, #"Don't know"
      is.na(INDFMMPC) ~ NA_character_  
    )
  )




########### Alcohol  ###########
data <- data %>%
  mutate(
    ALQ121 = if_else(RIDAGEYR < 18, 0, ALQ121),  # Assign 0 for age < 18
    alcohol_frequency = case_when(
      ALQ121 == 0 ~ '0', #"Never in the last year"
      ALQ121 == 1 ~ '10', #"Every day"
      ALQ121 == 2 ~ '9', #Nearly every day"
      ALQ121 == 3 ~ '8', #"3 to 4 times a week"
      ALQ121 == 4 ~ '7', #"2 times a week"
      ALQ121 == 5 ~ '6', #Once a week
      ALQ121 == 6 ~ '5', #2 to 3 times a month
      ALQ121 == 7 ~ '4', #Once a month
      ALQ121 == 8 ~ '3', #7 to 11 times in the last year
      ALQ121 == 9 ~ '2', #3 to 6 times in the last year
      ALQ121 == 10 ~ '1', #1 to 2 times in the last year
      ALQ121 == 77 ~ NA_character_, #"Refused"
      ALQ121 == 99 ~ NA_character_, #"Don't know"
      is.na(ALQ121) ~ NA_character_
    )
  )

########### Diabete ###########
data <- data %>%
  mutate(
    diabetes_status = case_when(
      DIQ010 == 1 ~ '1', #Yes
      DIQ010 == 2 ~ '0', #No
      DIQ010 == 3 ~ '0.5', #Borderline
      DIQ010 == 7 ~ NA_character_, # "Refused"
      DIQ010 == 9 ~ NA_character_, # "Don't know"
      is.na(DIQ010) ~ NA_character_  # Handle missing values
    ))



########### Second hand smoke ########### 
data <- data %>%
  mutate(
    household_smoking_status = case_when(
      SMD460 == 0 ~ '0', #No one in household is a smoker
      SMD460 == 1 ~ '1', #1 household member is a smoker
      SMD460 == 2 ~ '2', #2 or more household members are smokers
      SMD460 == 777 ~ NA_character_, #Refused
      SMD460 == 999 ~ NA_character_, #"Don't know"
      is.na(SMD460) ~ NA_character_  # Handle missing values
    )
  )



########### Fast food ########### 
# add the range of values together, from 0 - 90, if added value over 90, or either one is 6666, mark it as 90+, if either one is 7777 or 9999 and the other is 0 - 90 or 6666, use the other value. If both are 7777 or 9999, make as 7777 or 9999 correspsoned

data <- data %>%
  rowwise() %>%
  mutate(
    fast_food_consumption = case_when(
      # If either value is 6666 (More than 90 times), mark as "99"
      DBD905 == 6666 | DBD910 == 6666 ~ "99",
      
      # If the sum of values exceeds 90, mark as "99"
      sum(c(DBD905, DBD910), na.rm = TRUE) > 90 ~ "99",
      
      # If one value is 7777 or 9999 and the other is 0-90 or 6666, use the valid value
      DBD905 %in% c(7777, 9999) & DBD910 %in% c(0:90, 6666) ~ as.character(DBD910),
      DBD910 %in% c(7777, 9999) & DBD905 %in% c(0:90, 6666) ~ as.character(DBD905),
      
      # If both are 7777 or 9999, retain the corresponding value
      DBD905 %in% c(7777, 9999) & DBD910 %in% c(7777, 9999) ~ as.character(max(DBD905, DBD910)),
      
      # Otherwise, add the values (if both are 0-90) and return the sum
      TRUE ~ as.character(sum(c(DBD905, DBD910), na.rm = TRUE))
    )
  ) %>%
  ungroup()


########### sleep time ########### 
# Convert time to decimal hours
time_to_decimal <- function(time) {
  ifelse(
    time %in% c("77777", "99999") | is.na(time),
    NA,
    as.numeric(hms(paste0(time, ":00"))) / 3600
  )
}

data <- data %>%
  mutate(
    # Convert sleep and wake times to decimal hours
    sleep_weekdays = time_to_decimal(SLQ300),
    wake_weekdays = time_to_decimal(SLQ310),
    sleep_weekends = time_to_decimal(SLQ320),
    wake_weekends = time_to_decimal(SLQ330),
    
    # Calculate sleep hours, accounting for crossing midnight
    sleep_hours_weekdays = case_when(
      !is.na(sleep_weekdays) & !is.na(wake_weekdays) ~ 
        ifelse(wake_weekdays < sleep_weekdays, wake_weekdays + 24 - sleep_weekdays, wake_weekdays - sleep_weekdays),
      TRUE ~ NA_real_
    ),
    sleep_hours_weekends = case_when(
      !is.na(sleep_weekends) & !is.na(wake_weekends) ~ 
        ifelse(wake_weekends < sleep_weekends, wake_weekends + 24 - sleep_weekends, wake_weekends - sleep_weekends),
      TRUE ~ NA_real_
    ),
    
    # Calculate weighted sleep directly
    sleep_hours = (5 * sleep_hours_weekdays + 2 * sleep_hours_weekends) / 7
  )

```

## check age distribution

```{r}
data %>% ggplot(aes(x = RIDAGEYR)) +
  geom_histogram(binwidth = 10, fill = "blue", color = "black", alpha = 0.7) +
  labs(
    title = "Histogram of Weighted Sleep Hours",
    x = "Weighted Sleep Hours",
    y = "Frequency"
  ) +
  theme_minimal()
```


## AGG data
```{r}
agg_data <- data %>%
  select(BMXBMI, bmi_category, RIDAGEYR, milk_type, milk_consumption, poverty_category, alcohol_frequency, diabetes_status, household_smoking_status, fast_food_consumption, sleep_hours) %>%
  rename(
    bmi = BMXBMI,
    age = RIDAGEYR
  )

head(agg_data)
  
```

# Exploratory Data Analysis
```{r}
# Summary of data
summary(agg_data)

# Check data structure
str(agg_data)
```

## Missingness
NA Summary
```{r}
library(knitr)

# Create a summary table of missing values
na_summary <- data.frame(
  Variable = colnames(agg_data),
  NA_Count = colSums(is.na(agg_data)),
  NA_Percentage = round(colSums(is.na(agg_data)) / nrow(agg_data) * 100, 2)
)

# Pretty table using kable
kable(na_summary, format = "markdown", col.names = c("Variable", "NA Count", "NA Percentage (%)"))
```

```{r}
final_data <- agg_data %>%
  filter(!is.na(milk_type), !is.na(milk_consumption), !is.na(poverty_category), !is.na(alcohol_frequency), !is.na(diabetes_status), !is.na(household_smoking_status), !is.na(fast_food_consumption), !is.na(sleep_hours))
```

```{r}
# Summary of data
summary(final_data)

# Check data structure
str(final_data)
```
## factorize Ordinal Variables 
predictors are ordinal, encoding them to reflect their order.
```{r}
final_data <- final_data %>%
  mutate(bmi_category = factor(bmi_category, 
      levels = c("Underweight", "Normal", "Overweight", "Obesity"), 
      ordered = TRUE)) %>%
  mutate(milk_type = factor(
      milk_type, 
      levels = c("0", "1", "2", "3", '4', '8', '9'), 
      ordered = TRUE)) %>%
  mutate(milk_consumption = factor(
      milk_consumption, 
      levels = c("0", "1", "2", "3", '9'), 
      ordered = TRUE))

final_data <- final_data %>%
  mutate(poverty_category = as.numeric(poverty_category)) %>%
  mutate(alcohol_frequency = as.numeric(alcohol_frequency)) %>%
  mutate(diabetes_status = as.numeric(diabetes_status)) %>%
  mutate(household_smoking_status = as.numeric(household_smoking_status)) %>%
  mutate(fast_food_consumption = as.numeric(fast_food_consumption))
```


## realtionship between variable of interest
```{r, fig.width=10, fig.height=5}
custom_labels <- c(
  "Underweight" = "Under\nWeight",
  "Normal" = "Normal",
  "Overweight" = "Over\nWeight",
  "Obesity" = "Obesity"
)

facet_labels <- c(
  "poverty_category" = "Poverty Category",
  "alcohol_frequency" = "Alcohol Frequency",
  "diabetes_status" = "Diabetes Status",
  "household_smoking_status" = "Household Smoking Status",
  "fast_food_consumption" = "Fast Food Consumption",
  "sleep_hours" = "Hours of Sleep",
  "milk_type" = "Milk Type",
  "milk_consumption" = "Milk Consumption"
)


plot <- final_data %>%
  mutate(
    milk_type = as.numeric(as.character(milk_type)),
    milk_consumption = as.numeric(as.character(milk_consumption))
  ) %>%
  pivot_longer(
    cols = c(poverty_category, alcohol_frequency, diabetes_status, household_smoking_status, fast_food_consumption, sleep_hours, milk_type, milk_consumption),
    names_to = "variable",
    values_to = "value"
  ) %>%
  ggplot(aes(x = bmi_category, y = value, fill = bmi_category)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red")+
  scale_x_discrete(labels = custom_labels) +
  facet_wrap(~ variable, scales = "free_y", labeller = labeller(variable = facet_labels), ncol = 4) +
  labs(
    title = "Relationship Between Variables of Interest and BMI Categories",
    x = "BMI Category",
    y = "Variable of Interest Value"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
    strip.text = element_text(size = 12), # larger facet title
    axis.title = element_text(size = 12, face = "bold")
        ) 
  ggsave("relation_btw_variable.png", plot=plot, width = 10, height = 5)
```

- no obvious outliers

## correlation
- ordinal variable
- non-linear realtionship 
-> use spearman correlation
```{r}
cor_matrix <- final_data %>%
  select(poverty_category, alcohol_frequency, diabetes_status, household_smoking_status, fast_food_consumption, sleep_hours, milk_type, milk_consumption) %>%
  mutate(
    milk_type = as.numeric(as.character(milk_type)),
    milk_consumption = as.numeric(as.character(milk_consumption))
  ) %>% 
  cor(method = "spearman")

# Visualize the correlation matrix

cor_plot <- ggcorrplot(cor_matrix, method = "circle", title = "Correlation Between Predictors")
ggsave("correlation_matrix.png", plot=cor_plot)
```
```{r}
predictors <- final_data %>%
  select(poverty_category, alcohol_frequency, diabetes_status, household_smoking_status, fast_food_consumption, sleep_hours, milk_type, milk_consumption) %>%
  mutate(
    milk_type = as.numeric(as.character(milk_type)),
    milk_consumption = as.numeric(as.character(milk_consumption))
  )

results <- data.frame(
  Var1 = character(),  # First variable
  Var2 = character(),  # Second variable
  Correlation = numeric(),
  T_Statistic = numeric(),
  P_Value = numeric(),
  stringsAsFactors = FALSE
)

for (i in 1:(ncol(predictors) - 1)) {
  for (j in (i + 1):ncol(predictors)) {
    # Get variable names
    var1 <- colnames(predictors)[i]
    var2 <- colnames(predictors)[j]
    
    # Perform correlation test
    test <- cor.test(predictors[[var1]], predictors[[var2]], method = "pearson")  # Use "spearman" if needed
    
    # Extract statistics
    correlation <- test$estimate
    t_statistic <- test$statistic
    p_value <- test$p.value
    
    # Store results in dataframe
    results <- rbind(results, data.frame(
      Var1 = var1,
      Var2 = var2,
      Correlation = correlation,
      T_Statistic = t_statistic,
      P_Value = p_value
    ))
  }
}

```

```{r}

results %>%
  select(Var1, Var2, T_Statistic, P_Value) %>%
  rowwise() %>%
  mutate(Output = paste("For pair (", Var1, ", ", Var2, "): T-Statistic = ", T_Statistic, ", P-Value = ", P_Value, sep = "")) %>%
  pull(Output) %>%
  writeLines()
```


```{r}
results$Bonferroni_Adjusted_P <- p.adjust(results$P_Value, method = "bonferroni")

results %>%
  select(Var1, Var2, T_Statistic, Bonferroni_Adjusted_P) %>%
  rowwise() %>%
  mutate(Output = paste("For pair (", Var1, ", ", Var2, "): T-Statistic = ", T_Statistic, ", Padj-Value = ", Bonferroni_Adjusted_P, sep = "")) %>%
  pull(Output) %>%
  writeLines()
```
## scale
depend on model choosen

## Transformation
depend on model choosen



















